Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-13 sftrends_manu.Rnw
dinc10 figure description donein text

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.  Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between the two is shown in \cref{fig:dinc10}c such that years with model predictions greater or less than the flow-normalized values correspond wiht long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.  Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range

Emphasize the information the model provides relative to the observed time series.  A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.  The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year. Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).  The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.  Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).  Emphasize the summer/fall change in the 2000s, why is this?  Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN? Relate to conceptual diagram.

Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.

>>>>

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.  Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.  Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.  Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s. % The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}

A comparison of trends with flow-normalized components relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.  These comparisons are made relative to changes in the magnitude, significance, and direction of the trend, all of which have important implications for decision-making. Overall, the differences in apparent trends underscore the critical importance of considering flow effects in decision-making, where flow is affected by complex relationships between inflow, water withdrawals, and exchange with the larger Bay.
